Bird Audio Detection Challenge

### 2017-03-17 paper/main.tex
Polished paper a bit

\newenvironment{alt_enumerate}
{ \begin{enumerate}
		\setlength{\itemsep}{1pt}
		\setlength{\parskip}{1pt}
		\setlength{\parsep}{1pt} }
	{ \end{itemize} }

@@

@@

	Acoustic detection of birds in the wild is useful in many biological
	scenarios, including estimation of bird population and species-specific
	acoustic monitoring. Most of the current methods are either manual or
	semiautomatic, requiring labor-intensive processing/tuning. We hereby
	present three alternative methods, as part of the Bird Audio Detection
	Challenge, hosted by Queen Mary University of London. Two methods are of
	our own submission, with the third one as the winning solution submitted
	by Thomas Grill. All methods utilize deep learning for automatic,
	tuning-free bird presence detection.

@@

Acoustic detection of birds in the wild is useful in many biological
scenarios, including estimation of bird population and species-specific
acoustic monitoring. Most of the current methods are either manual or
semiautomatic, requireing labor-intensive processing/tuning.

In response to the need for automatic, tuning-free bird presentce detection
algorithm, Bird Audio Detection Challenge was created, hosted by Queen Mary
University of London and other universities. Three datasets were released
as part of the challenge, and teams across the world compete for the best
performance on the held-out dataset.

In section 2, we will discuss some of the competition details, including
the datasets used, task description and the evaluation metrics. Section 3
will present three methods for bird acoustic detection, two of which are
our submission to the challenge, and the third one being the best
performing method submitted by Thomas Grill. Section 4 will discuss the
evaluation results of the three methods, and Section 5 will present future
possible work based off of this challenge.

@@

For the challenge, teams have to design a system that takes a short audio
clip, and returns a binary label denoting the presence/absence of bird
sound of any kind. Teams are also encouraged to output a continuous,
probability score in the range of 0 to 1 for evaluation purposes.

@@

\textbf{freefield1010:} 7000 excerpts from field recordings around the
world, gathered by the FreeSound project.

\textbf{Warblr:} Crowdsourced project with many smartphone recordings from
around the United Kingdom. Very diverse in sound categories, even including
human bird imitations.

\textbf{Chernobyl:} Dataset created as part of Dr Wood's research into the
long-term effects of the Chernobyl accident on local ecology. Approximately
10,000 hours to date

Each dataset contains audio clips of roughly 10 seconds. For the challenge,
all datasets' audio files are provided, but only \textbf{freefield} and
\textbf{Warblr}'s labels are given. Held-out dataset consists primarily of
the \textbf{Chernobyl} dataset, along with some from the other two.

@@

%-------------------------------------------------------------------------
\section{Bird Acoustic Detection}

@@

All audio clips are 10 seconds, mono at 44100 Hz, which gives us 441000
frames of audio. We divide all frames by 32768 to convert values of
individual frames to between -1 and 1, and randomly crop out 400000 frames
as input to our network.

@@

Our first method resembles basic characteristics of SoundNet and residual
net. The network is fully convolutional, with one-dimensional convolutions
followed by nonlinearities (ReLU) layers. This network is suitable for
audio signals, as it is invariant to translation and allows stacking of
convolutional layers for detecting high-level features.

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=0.5\textwidth]{v5_diagram}
	\caption{Network v5 Architecture. Content with box refers to kernel size,
	number of filters, and stride size}
\end{figure}

@@

Network can be divided into three stages. In the first stage, we first
create two copies of the input, one reshaped to 100000 by 4, and the other
reshaped to 200000 by 2. \(x_{100000*4}\) is then passed to a convolutional
layer with kernel size \([3,4]\) while \(x_{200000*2}\) is passed to a
series of 5 pooling layers, followed by a convolutional layer at the end of
each pooling layer. Note here that the input of one pooling layer is the
output of the previous pooling layer, the outputs of the convolutional
layers serve as features extracted at audio of different resolution.

In the second stage, we combined the outputs of the 6 convolutional layers
in Stage 1 into one feature vector. First, we find the maximum for each
convolutional output along the thrid dimension. We denote these as the
\textit{energies} for each convolutional layer. We then concatenate them
with their corresponding convolutional output, run each concatenated
feature vector through a pooling layer of specified size, and concatenate
all 6 feature vectors into one along the third dimension. Due to different
lengths of output for the 6 convolutional layers, we pass them through
pooling layers of different sizes to ensure that they are of the same
dimension for concatenation.

In the final stage, we pass the output from Stage 2 through 7 additional
convolutional layers, with a \(max()\) operation between the 4th and 5th
layer. Final output is then flattened into size (1,2), ready to be used in
a softmax classification loss function.

@@

This method is the winning algorithm of the challenge, submitted by Thomas
Grill. The neural network used consists of 4 convolutional layers with
pooling, and 2 dense layers. Multiple signal pre-processing techniques were
used, including silence/noise trimming, and a different data augmentation
technique involving shifting the spectra in time and frequency.
Pseudo-labeling and bagging techniques are also deployed, which we will
discuss in section 4.

@@

During our development and training of the network, we experimented with
various pre- and postprocessing techniques, as well as some training
strategies.

@@

The given labeled data from both dataset is split into 10 folds. Within
each fold, number of positive and negative samples are balanced. During
training, number of samples loaded from each dataset are also balanced, to
ensure no bias.

For in-house testing, 9 folds are used for training and 1 folf for testing.
For submission, all 10 folds are used for training.

For our method 1 and 2, we implemented them in python with Tensorflow
machine learning library. We trained from scratch, with learning rate of
0.03, learning rate decay by a factor of 0.1 every 40000 iterations, and a
total of 300000 iterations.

@@

We deployed data augmentation as our only pre-processing technique. Given
one base sound clip, positive or negative, and another negative sound clip,
we concatenate them along the y-axis, with the base clip weighted by a
factor \(r\) between 0.25 and 1, and the negative sound clip weighted by a
factor \(1-r\). So we have:
\begin{equation}
	x_{augmented}=r*x_{signal} + (1-r)*x_{negative}
\end{equation}
If \(r\) equals to 1, then there would be no augmentation, and if \(r\)
equals to 0.25, only 25\% of the base sound clip will be preserved. We
prevent \(r\) from going below 0.25 to ensure bird sound is still
detectable after augmentation.

Label for the augmented clip will remain the same as the base clip, as the
augmentation is an \(|\) process, preserving the original label when \(|\)
with a negative label.

@@

One main difficulty we faced is the small size of training data. With no
regularization, our networks consistently over-fit to the training data,
perform near perfect on training and our own hold-out test data, but
perform poorly on the challenge data. To compensate for the lack of data,
we reduce the capacity of our models by reducing the number of filters for
each convolutional layer. We do so until the network cannot perform
perfectly on the training and testing data.

@@

This post-processing technique averages prediction scores from many models.
Due to the random cropping of audio signals during the data loading phase,
different evaluation of the same audio clip will yield slightly different
results. Therefore we predict 100 sets of scores for the held-out challenge
dataset, and average them to obtain one set of scores that was submitted.

@@

This training strategy addresses the issue of dataset bias. It is apparent
that the three given datasets have disjoint domains. In addition, within
each dataset, data is skewed either towards mostly positive samples or
mostly negative ones. During training, it is very likely that the network
will cheat by identifying which dataset the clip is from, and make its
prediction based off of that information. In response, we take the audio
features near the end of the network, pass it through another convolutional
layer, and...

@@

\begin{alt_enumerate}
	\item{Train a model with only labeled data}
	\item{Predict scores of the held-out challenge dataset}
	\item{Find threshold where within it, predictions are accurate enough to
		be used as extra labels}
	\item{Add all predictions within threshold, and retrain or fine-tune
		network}
\end{alt_enumerate}

@@

For method 1 and 2, no signal pre-processing techniques were deployed,
resulting in very noisy inputs to our network. It will be interesting to
find out if pre-processing will aid the prediction accuracy of our network.

@@

Bird acoustic detection is an important task for many biological
applications, and the challenge has advanced this field in many ways. We
presented three methods, first two with a focus on unique network
architecture, and the third one as the embodiment of many years of machine
listening research by Thomas Grill. Many topics are still open for
research, like bird species identification and detection.

>>>>

\newenvironment{alt_enumerate}
{ \begin{enumerate}
		\setlength{\itemsep}{1pt}
		\setlength{\parskip}{1pt}
		\setlength{\parsep}{1pt} }
	{ \end{enumerate} }

@@

% TODO: Add Tawfiq in without messing up format

@@

	Acoustic detection of birds in the wild is useful in many biological
	scenarios, including estimation of bird population and species-specific
	acoustic monitoring. Most of current methods are either manual or
	semi-automatic, requiring labor-intensive processing/tuning. We hereby
	present three alternative methods, as part of the Bird Audio Detection
	Challenge, hosted by Queen Mary University of London. Two methods are our
	own submission, with the third one as the winning solution submitted by
	Thomas Grill. All methods utilize deep neural networks for automatic,
	tuning-free bird presence detection.

@@

Much biological research relies on bird acoustic detection in audio signals,
as indicated by \cite{bad}. Over the recent years, a large amount of audio
data is recorded, but the bird detection methods are often manual or
semi-automatic, requiring intensive labor or tuning.

In response to the need for automatic, tuning-free bird presentce detection
algorithm, \textit{Bird Audio Detection Challenge} was created, hosted by
Queen Mary University of London and other universities. Three datasets were
released as part of the challenge, and teams across the world compete for
the best performance on the task of bird audio detection.

In section 2, we discuss some of the competition details, including the
datasets used, task description and the evaluation metrics. Section 3
presents three methods for bird acoustic detection, two of which are our
submission to the challenge, and the third one being the best performing
method submitted by Thomas Grill. Section 4 discusses some experiments and
techniques used to improve the peformance. Section 5 displays the
evaluation results of the three methods, and section 6 presents future
possible work with this challenge.

@@

For the challenge, teams have to design a system that takes a short audio
clip, and returns a binary label denoting the presence/absence of bird
sound of any species. Teams are also encouraged to output a continuous,
probability score in the range of 0 to 1 for evaluation purposes.

@@

% TODO: Should FreeSound be italicized?

\textbf{freefield1010:} 7,000 excerpts from field recordings around the
world, gathered by the FreeSound project.

\textbf{Warblr:} Crowd-sourced project with many smartphone recordings from
around the United Kingdom. Very diverse in sound categories, even including
human bird imitations.

% TODO: Should link to project be included?

\textbf{Chernobyl:} Part of TREE (Transfer-Exposure-Effects) research
project. Consists of audio in the Chernobyl Exclusion Zone.

Each dataset contains audio clips of roughly 10 seconds. For the challenge,
all datasets' audio files are provided, but only \textbf{freefield} and
\textbf{Warblr}'s labels are given. Held-out dataset consists primarily of
the \textbf{Chernobyl} dataset, along with some from \textbf{freefield} and
\textbf{Warblr}.

@@

%-------------------------------------------------------------------------
\section{Methodology}

@@

All individual audio clips are 10 seconds, mono at 44,100 Hz, which gives us
441,000 frames of audio. We divide all frames by 32,768 to convert values of
individual frames to between -1 and 1, and randomly crop out 400,000
continuous frames, resulting in a one-dimensional vector with length 400,000
as our input to the network.

@@

Our first method resembles basic characteristics of SoundNet by Aytar et
al.\cite{soundnet} and Residual Net by He et al.\cite{residual_net}. The
network is fully convolutional, with one-dimensional convolutions followed
by nonlinearities (ReLU) layers.  This network is suitable for audio
signals, as it is invariant to translation and allows stacking of
convolutional layers for detecting high-level features.

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=0.5\textwidth]{v5_diagram}
	\caption{Network v5 Architecture. Content within convolutional blocks
	refers to kernel size, number of filters, and stride size}
\end{figure}

@@

Network can be divided into three stages. In the first stage, we first
create two copies of the input, one reshaped to 100,000 by 4, and the other
reshaped to 200,000 by 2. \(x_{100,000*4}\) is then passed through a single
convolutional layer while \(x_{200,000*2}\) is passed through a series of 5
pooling layers, followed by a convolutional layer at the end of each
pooling layer. Note here that the input of one pooling layer is the output
of the previous pooling layer, the outputs of the convolutional layers
serve as features extracted at input of different resolution.

In the second stage, we combined the outputs of the 6 convolutional layers
in Stage 1 into one feature vector. First, we find the maximum for each
convolutional output. We denote them as the energy for each convolutional
layer. We then concatenate them with their corresponding convolutional
output, pass each concatenated feature vector through a pooling layer of
specified size, and concatenate all 6 feature vectors into one. Due to
different lengths of output for the 6 convolutional layers in stage 1, we
pass them through pooling layers of different sizes to ensure that they are
of the same dimension for concatenation.

In the final stage, we pass the output from Stage 2 through a batch
normalization layer, followed by 7 additional convolutional layers, with a
\(max()\) operation between the 4th and 5th layer. Final output is then
flattened into size (1,2), ready to be used in a softmax classification
loss function.

@@

This method is the winning algorithm to the challenge, submitted by Thomas
Grill. The neural network uses processed spectrograms of audio clips as
input, instead of a one-dimensional vector of signals.The network consists
of 4 convolutional layers with pooling, and 2 dense layers. Multiple signal
pre-processing techniques are used, including silence/noise trimming, and a
different data augmentation technique involving shifting the spectrograms
in time and frequency.  Pseudo-labeling and bagging techniques are also
deployed, which we will discuss in section 4.

@@

During our development, we experimented with various pre- and
postprocessing techniques, as well as some training strategies.

@@

The given labeled data from both datasets is split into 10 folds. Within
each fold, number of positive and negative samples are balanced. During
training, number of samples loaded from each datasets are also balanced, to
ensure no bias.

For in-house testing, 9 folds are used for training and 1 fold for testing.
For submission, all 10 folds are used for training.

For method 1 and 2, we implemented them in python with Tensorflow machine
learning library. We trained from scratch, with learning rate of 0.03,
learning rate decay by a factor of 0.1 every 40,000 iterations, and a total
of 300,000 iterations.

For method 3, code base is available online at...

% TODO: How to include code base link?

@@

We deploy data augmentation as our only pre-processing technique for method
1 and 2. Given one base sound clip, positive or negative (bird
present/absent), and another negative sound clip, we concatenate them along
the y-axis, with the base clip weighted by a factor \(r\) between 0.25 and
1, and the negative sound clip weighted by a factor \(1-r\). So we have:
\begin{equation}
	x_{augmented}=r*x_{base} + (1-r)*x_{negative}
\end{equation}
If \(r\) equals to 1, then there would be no augmentation, and if \(r\)
equals to 0.25, only 25\% of the base sound clip will be preserved. We
sample \(r\) on a positively skewed interval (0.25,1) to ensure most bird
sounds in augmented clips are still detectable.

Label for the augmented clip will remain the same as the base clip, as the
augmentation is an \textit{or} process, preserving the original label when 
\textit{or} with a negative label.

@@

One main difficulty we faced is the small size of training data. With no
regularization, our networks consistently over-fit to the training data,
perform near perfect on training and our own held-out test data, but
perform poorly on the challenge data. To compensate this, we reduce the
capacity of our models by reducing the number of filters for each
convolutional layer. We do so until the network cannot perform perfectly on
the training and testing data.

@@

This post-processing technique averages prediction scores from many
sessions. Due to the random cropping of audio signals during the data
loading phase, different evaluation of the same audio clip will yield
slightly different results. Therefore we predict 100 sets of scores for the
held-out challenge dataset, and average them to obtain one set of scores
that was submitted.

For method 3, instead of averaging sets of output scores from the same
model, different models' outputs were averaged. Please refer to Grill's
code base for more details.

@@

This training strategy addresses the issue of dataset bias. It is apparent
that the three given datasets have disjoint sound domains. In addition,
within each dataset, data is skewed either towards mostly positive samples
or mostly negative ones. During training, it is very likely that the
network will cheat by identifying which dataset the clip is from, and make
its prediction based off of that information. In response, we take the
audio features near the end of the network, pass it through another
convolutional layer, and...

@@

\begin{alt_enumerate}
	\item{Train a model with only labeled data}
	\item{Predict scores of the held-out challenge dataset}
	\item{Find threshold \(r\) such that predictions within range\\
		\(0<r \cup (1-r)<1\) are accurate enough}
	\item{Add all predictions within threshold, and retrain or fine-tune
		network}
\end{alt_enumerate}

\(r\) can be obtained through manual spot-checking of the predictions. Too
much time spend on finding the threshold will defeat the purpose of the
challenge.

@@

For method 1 and 2, no signal pre-processing techniques were deployed,
resulting in very noisy inputs to our network. It will be interesting to
find out if pre-processing will aid the prediction accuracy of our two
networks.

@@

Bird acoustic detection is an important task for many biological
applications, but the massive amount of data available makes it difficult
to extract any useful information from it. The Bird Audio Detection
challenge has advanced this field in many ways, and many excellent
automatic, tuning-free bird acoustic detection algorithms are introduced.
We presented three methods, first two with a focus on unique network
architecture, and the third one as the embodiment of many years of machine
listening research.  Many topics are still open for research, including
many tasks that already have a solution, but only achieved through manual
or labor intensive.
